Interpretable Neural Networks using EAGGA

### 2025-03-14 paper/main.tex
HPO model-free EA intro

% model free
% grid + random search
The most basic model-free strategies for blackbox optimization are grid and random search.
In the latter, the user defines a range of interest for each HP and then evaluating points along a grid within the cartesian product of those.
This has the drawback of scaling extremely poorly in both number of HP dimensions and number of query points per range of interest.
Random search, on the other hand, randomly samples a value for each HP until it runs out of budget. Its explorative nature, ease of use, and no assumptions about the model
makes it a good baseline. It is also superior to grid search in cases were one or more HPs have little impact on model performance, as for a given budget $B$,
each HP will likely be queried with $B$ different values, whereas for grid search each HP will only be queried with $B^{1/N}$ different values for $N$ HPs.

(evolutionary algorithms)
- another model-free class of optimisation algorithms, inspired by naturally occuring evolution
- based on a population, which iteratively (iteration = generation) generates $\lambda$ offspring via 3 operators (all sub-bullets from \citet[pp. 10-]{genetic_algos})
  * \textbf{reproduction}: pick an individual to reproduce with prob proportional to its fitness
  * \textbf{crossover}
    - select 2 ``parents'' from pool of reproducing individuals
    - for a position $k$ in their HP configs, with certain prob, swap their values after position $k$, yielding 2 new children
    % TODO: use example from book page 12
  * \textbf{mutation}: with certain prob, change values of an individual, e.g. by adding Gaussian noise (Gaussian mutation) to real values, or flipping bit for binary values
- selection at end of each genertion: keep $\mu$ best individuals (based on some fitness function), either from only offspring (``$(\mu,\lambda)$-selection'')
  or (more commonly) from population + offspring (``$(\mu+\lambda)$-selection'', guaranteed to keep best individual)
- pro: conceptually simple + can handle even complex parameter spaces (continuous, discrete, hierarchical, etc.) given appropriate implementation of operators
all above starting at HPO from \citep[chap. 1.3]{feurer_hyperparameter_2019}
- prominent examples
  * CMA-ES
    - ``Covariance Matrix Adapation Evolution Strategy''
    - offspring generation exclusively via multivariate normal \citep[p. 8]{hansen2023cmaevolutionstrategytutorial}
      * mean = weighted average of previous generation
      * covariance = weighted covariance of previous generation, with weighting as for mean
    - weighing scheme done to sample in a way as to reproduce previously successful (i.e. selected) steps \citep[p. 11]{hansen2023cmaevolutionstrategytutorial}
  * differential evolution \citep[-]{differential_evolution}
    - init population randomly so that entire param space is covered
      * $D$-dim vectors $\boldsymbol{x}_{i,G}$ with $i=1,2,...,n$ for gen $G$
    - mutation
      * for each $\boldsymbol{x}_{i,G}$, generate \textbf{mutant vector} $\boldsymbol{v}_{i,G+1}=\boldsymbol{x}_{r_1,G}+F\cdot(\boldsymbol{x}_{r_2,G}-\boldsymbol{x}_{r_3,G})$
      * $r_1,r_2,r_3\in\{1,2,...,n\}$ mutually different random idx + different from $i$
      * constant $F\in[0,2]$
    - crossover
      * \textbf{trial vector} $\boldsymbol{u}_{i,G+1}=(u_{i,G+1}^{(1)},u_{i,G+1}^{(2)},...,u_{i,G+1}^{(D)})$
        - chose random index $R\in\{1,2,...,D\}$
        - sample $p\sim U[0,1]$
        - define crossover constant $\text{CR}\in[0,1]$
        - $u_{i,G+1} = \begin{dcases}
          v_{i,G+1}^{(j)} & \text{if } p\le\text{CR}\text{ or } j=R \\
          x_{i,G}^{(j)} & \text{else}
         \end{dcases}$
    - selection: greedy, compare fitness of $\boldsymbol{u}_{i,G+1}$ with $\boldsymbol{x}_{i,G}$, pick better

>>>>

% model free
% grid + random search
The most basic model-free strategies for blackbox optimization are grid and random search.
In the latter, the user defines a range of interest for each HP and then evaluating points along a grid within the cartesian product of those.
This has the drawback of scaling extremely poorly in both number of HP dimensions and number of query points per range of interest.
Random search, on the other hand, randomly samples a value for each HP until it runs out of budget. Its explorative nature, ease of use, and no assumptions
about the model makes it a good baseline. It is also superior to grid search in cases were one or more HPs have little impact on model performance,
as for a given budget $B$, each HP will likely be queried with $B$ different values, whereas for grid search each HP will only be queried with $B^{1/N}$
different values for $N$ HPs. \citep[chap. 1.3]{feurer_hyperparameter_2019}

% evolutionary algorithms
Another popular class of model-free optimizers are evolutionary algorithms (EAs), which are conceptually simple and can handle even complex parameter spaces,
given appropriate implementation of operators.
EAs are iteratively evolving a population of individuals (an individual is simply a hyperparameter configuration) of size $\mu$, where in each iteration (generation),
$\lambda$\footnote{We distinguish between $\lambda\in\mathbb{N}$ when referring to the offspring size of EAs and $\boldsymbol\lambda\in\boldsymbol\Lambda$ when
referring to a hyperparameter configuration.} offspring are generated via three operators \citep[pp. 10-TODO]{genetic_algos}.\\
\textbf{Reproduction} samples an individual from the population to reproduce with probability proportional to its fitnes. For HPO, fitness usually refers to the
performance metric the model resulting from the individual's HP configuration is evaluated on.\\
\textbf{Crossover} selects two ``parents'' from the pool of reproducing individuals. An index $k$ is sampled at random and all HP values from
index $k$ on are swapped, yielding two new ``children'' individuals.\\
\textbf{Mutation} randomly changes single HP values of an individual chosen for reproduction, e.g. by adding Gaussian noise to real values or flipping bits
on binary values.\\
At the end of each generation, the EA keeps the best $\mu$ individuals, either only from the offspring (``$(\mu,\lambda)$-selection'') or more commonly from
population and offspring (``$(\mu+\lambda)$-selection''), which is also called an ``elitist'' strategy, as it's guaranteed to keep the best individual.
\citep[chap. 1.3]{feurer_hyperparameter_2019}

%- another model-free class of optimisation algorithms, inspired by naturally occuring evolution
%- based on a population, which iteratively (iteration = generation) generates $\lambda$ offspring via 3 operators (all sub-bullets from \citet[pp. 10-]{genetic_algos})
%  * \textbf{reproduction}: pick an individual to reproduce with prob proportional to its fitness
%  * \textbf{crossover}
%    - select 2 ``parents'' from pool of reproducing individuals
%    - for a position $k$ in their HP configs, with certain prob, swap their values after position $k$, yielding 2 new children
    % TODO: use example from book page 12
%  * \textbf{mutation}: with certain prob, change values of an individual, e.g. by adding Gaussian noise (Gaussian mutation) to real values, or flipping bit for binary values
%- selection at end of each genertion: keep $\mu$ best individuals (based on some fitness function), either from only offspring (``$(\mu,\lambda)$-selection'')
%  or (more commonly) from population + offspring (``$(\mu+\lambda)$-selection'', guaranteed to keep best individual)
%- pro: conceptually simple + can handle even complex parameter spaces (continuous, discrete, hierarchical, etc.) given appropriate implementation of operators
- prominent examples
  * CMA-ES
    - ``Covariance Matrix Adapation Evolution Strategy''
    - offspring generation exclusively via multivariate normal \citep[p. 8]{hansen2023cmaevolutionstrategytutorial}
      * mean = weighted average of previous generation
      * covariance = weighted covariance of previous generation, with weighting as for mean
    - weighing scheme done to sample in a way as to reproduce previously successful (i.e. selected) steps \citep[p. 11]{hansen2023cmaevolutionstrategytutorial}
  * differential evolution \citep[-]{differential_evolution}
    - init population randomly so that entire param space is covered
      * $D$-dim vectors $\boldsymbol{x}_{i,G}$ with $i=1,2,...,n$ for gen $G$
    - mutation
      * for each $\boldsymbol{x}_{i,G}$, generate \textbf{mutant vector} $\boldsymbol{v}_{i,G+1}=\boldsymbol{x}_{r_1,G}+F\cdot(\boldsymbol{x}_{r_2,G}-\boldsymbol{x}_{r_3,G})$
      * $r_1,r_2,r_3\in\{1,2,...,n\}$ mutually different random idx + different from $i$
      * constant $F\in[0,2]$
    - crossover
      * \textbf{trial vector} $\boldsymbol{u}_{i,G+1}=(u_{i,G+1}^{(1)},u_{i,G+1}^{(2)},...,u_{i,G+1}^{(D)})$
        - chose random index $R\in\{1,2,...,D\}$
        - sample $p\sim U[0,1]$
        - define crossover constant $\text{CR}\in[0,1]$
        - $u_{i,G+1} = \begin{dcases}
          v_{i,G+1}^{(j)} & \text{if } p\le\text{CR}\text{ or } j=R \\
          x_{i,G}^{(j)} & \text{else}
         \end{dcases}$
    - selection: greedy, compare fitness of $\boldsymbol{u}_{i,G+1}$ with $\boldsymbol{x}_{i,G}$, pick better
